k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-01-31 paper.Rmd
abtract and outline

    When assigning stimuli to sets in experimental psychology, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually substandard when done manually by researchers, automated algorithmic approaches are preferred to accomplish this goal. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) that simultaneously maximizes the diversity (corresponding to between-set similarity) and the dispersion (corresponding to pairwise stimulus dissimilarity within-sets). We provide a first independent assessment of BILS and present BILS-E, which extends BILS in three ways. First, a globally optimal value of dispersion is always returned. Second, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, for example to include cannot-link constraints in the anticlustering process. Third, BILS-E allows that other anticlustering objectives of between-group similarity, such as k-means or k-plus, can be optimized instead of only the diversity. Using our optimal approach as benchmark in a simulation study, we show that the original BILS identifies optimal dispersion values for much larger data sets than previously analyzed, but also highlight conditions where it fails to do so. A practical example shows how BILS-E can be used for stimulus assignment in experimental psychology. The original BILS as well as our extension are both freely available via the open source R package anticlust.

@@

>>>>

    When assigning stimuli to sets in experimental psychology, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) that simultaneously maximizes the diversity (corresponding to between-set similarity) and the dispersion (corresponding to pairwise stimulus dissimilarity within sets). We provide a first independent assessment of BILS and extend it in three ways. First, by prefixing an integer linear programming algorithm, our extension ensures that a globally optimal value of dispersion is always returned; without it, it is unknown if the BILS's result is optimal. Second, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, for example to prevent specific pairs of stimuli from appearing in the same condition. Third, our extension allows that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. Two simulation studies provide an extensive evaluation of the original BILS and our extension. Using the optimal approach as benchmark, we show that the original BILS identifies optimal dispersion values for much larger data sets than previously analyzed, but we also highlight conditions where it fails to do so reliably. A practical example shows how the BILS can be used for stimulus assignment in experimental psychology. The original BILS as well as our extension are both freely available via the open source R package anticlust.

@@

## Outline

- Start with use case: Stimulus assignment in within-subjects experiments (and stick with it throughout the paper, only briefly mention other use cases)
  * Needed: high similarity between stimulus set to reduce variance in responses attributed to other factors than the experimental manipulation. This is possible using `anticlust`, which has been applied many times since its introduction in @papenberg2020 (cite references)
  * Also needed (point of this paper!): High pairwise dissimilarity of stimuli in the same set. 
- Simultaneous optimization of between-set similarity and pairwise within-set dissimilarity was introduced by Brusco et al.'s alogrithm, which is also available in `anticlust` since 2021 (version 0.6.0)
- Here, we provide researchers with a practical introduction to their method and also present several extensions, which allow it to cover more use cases
  (1) Always return optimal solution
  (2) Use different distance matrix for dispersion and diversity (if different information is needed for both criteria, or to induce custom cannot-link constraints)
  (3) Extend the BILS's capacities towards other measures of between-group similarity (k-means / k-plus)
- Technical introduction to "vanilla" BILS
- Technical introduction to our extensions
  * Optimal dispersion method: Basic algorithm + graph coloring ILP
    * How to maintain optimal dispersion while the BILS runs? Outline three approaches!
  * Different distance matrices: does not change algorithm, just computation / implementation
  * Extend towards k-means: Use squared Euclidean distance (not really extension, just different input); compute "average" diversity if group sizes are unequal (k-plus is basically the same, but do not use the additional k-plus variables for the dispersion distance matrix)
- Simulation Study 1: Two research questions
  (1) How often does vanilla BILS find optimal solutions
  (2) Which of the three extension is best for optimizing between-group similarity while maintaining optimal dispersion?
- Simulation Study 2: More difficult conditions (do not continuous data; K > 4 groups)
- Application (presenting R code)
  * Optimize k-plus criterion using norming data; dispersion is based on pairwise orthographic dissimilarity between words
- Discussion

## Introduction

In within-subjects experiments, participants are presented with several stimulus sets that are affiliated with the different experimental conditions. Due to carry-over effects, showing the same stimuli in different conditions is oftentimes prohibited. This is evident in studies on memory, where studying a list of words in one conditions necessarily makes it impossible to repeat their study in a later phase; memory cannot be erased after an experimental block has been completed [@lahl2006]. While the experimental logic dictates that different lists must be used, the lists should be as parallel as possible with regard to variables that affect the response variable. @lintz2021 gave two reasons why it is important that study participants should be confronted with balanced lists: (a) if lists are not sufficiently parallel, the variance of within-subjects effect sizes due to the experimental manipulation will be inflated, reducing statistical power---if there is no proper statistical control of the stimulus materials, which is often the case [@judd2012]; (b) highly unbalanced sets may introduce secondary effects beyond the influence of the intrinsic stimulus features. That is, highly unequal lists may induce a strategic behavior shift between experimental blocks. Such a shift may for example result from suspicion, boredom or frustration, invalidating any observed differences between conditions. Importantly, this bias cannot be accounted for using statistical analysis.

**Using anticlustering to equate stimulus sets**

**BUT DISPERSION IS ALSO IMPORTANT!**

For example, in tasks of recognition memory, participants commonly have to decide if a target stimuli (e.g., a word) has been presented in an earlier phase of the experiment. In this case, words that are orthographically similar to a target word should serve as distractors.

The BILS algorithm has been available in the anticlust package since 2021, but we assume that many researchers do not yet know about its potential, also given that the introduction by @brusco2019 was rather technical in nature. For now, users seem to rely on the unicriterion methods focusing on between-set similarity only. Therefore, we provide readers with an application-focused introduction and extend BILS with parameters that may be useful when designing experiments.

**OLD**
